When the Pen Becomes a Sword: Retired Military Officers and Opinion-Editorial Usage in the US

### 2020-06-27 code/Submissionv1.Rmd
commit

---
title: "The Pen as a Sword: Retired Military Officers and Opinion-Editorial Usage"
author: "Pete Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper seeks to explain how retired military opinion-editorial usage has varied in the past 20 years. Using an original data set that captures more than 200 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, and the USA Today since 2000, I argue that recent opinion-editorial usage is weakening the norm of retired military officers remaining apolitical in two distinct ways. First, op-eds published in recent years by retired military officers address topics that are outside the scope of direct security or strategic concerns. Second, there is an increase in recently published op-eds which directly attack Presidential administrations, specifically that of Donald Trump. These two trends are concerning from the perspective of civil-military relations and raise further normative questions about how the American public receives information from military elites._

@@

In modern day America, the average citizen does not need to look or go far to hear or see an elite voice attempting to persuade them of the rightness or wrongness of certain courses of action. On the television and on the internet, "experts" in various forms take up a position on a certain issue and attempt to persuade, warn, and otherwise impact public opinion on a host of issues. To a considerable extent, this is a normal and everyday occurrence in a thriving democracy.

This paper examines the way in which a narrow subset of elites - retired military officers - has attempted to influence public opinion in contemporary America. It is an interesting question worthy of examination for several reasons and from a variety of perspectives. First, even after leaving active duty service, military officers remain part of an institution - the military - that maintains considerable trust among the American public. A 2018 Pew Research Center poll determined that 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. Therefore, when leaders of the most trusted institution in the country speak, the public seems to trust them to a high degree. Second, it is an interesting question because as an institution, the military prides itself on being apolitical. In an era of increasing political polarization [@layman_party_2006; @hare_polarization_2014], examining the extent to which opinion-editorial usage by retired military officers has changed may reveal one way in which the military institution has altered its behavior in a polarized political environment.

To be clear, I do not examine in this paper the efficacy of the opinion pieces authored by retired military officers. I do not address whether the actual op-ed pieces published by retired military officers actually sway public opinion. I am primarily concerned with what trends are evident in opinion-editorial publication from the perspective of the military as an institution: what variation is evident over the past 20 years, if any? Are there patterns that indicate deviation from historical norms and expectations of the military as an apolitical institution, and if so, what might this mean for the military?

This paper proceeds in three parts. First, I review the basis upon which the norm of apolitical behavior of the military as an institution is based, as well as describe how a polarized environment may impact these norms. Here I review academic literature on civil-military relations and public opinion, as well as current military doctrine. All three are relevant strands that inform our research question. Second, I describe the analytical approach and the data used in this project. Assembling an original data set that includes more than 200 opinion pieces published by retired military officers in major American newspapers since 2000, I conduct a mixed-methods approach. Finally, I describe my results as well as their implications. Overall, the data reveals that the vast majority of opinion pieces penned by retired military officers are in line with traditional civil-military norms. However, I find two trends that may reveal future problems for civil-military relations: first, retired military officers are writing about topics that are not immediately related to national security, and second, retired military officers are criticizing if not attacking political administrations more often than they did in previous eras.

# The Norm of Remaining Apolitical

As a norm, the military acting as an objective, apolitical institution is deep and storied in the United States. In a thriving democracy such as the US, an apolitical military can best serve the Republic by defending it regardless of which political party of group holds power. Unlike autocratic societies, where military coups are unfortunate yet frequent in history, the United States had not had a civil-military relations crisis that has risen to the point that a coup has been a real concern. This is not to say that that civil-military crises are infrequent in the US - indeed, they have been rather common - but that these crises have typically been resolved.

The political scientist Samuel Huntington, in his work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, argued for civil-military relations to take on a certain character. Written in 1957, during the Cold War, Huntington asserted that two distinct forces would always be in tension with one another in a liberal democracy. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced [@huntington_soldier_1957]. The second force was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society, and that these forces were at odds with each other. Huntington proceeded to argue that the key to resolving the tension between these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented plans to accomplish these goals [@huntington_soldier_1957]. To Huntington, separate operating spheres between the political leaders of a democracy and the military as an institution ensured that the military remained "professional," and thus, untainted by political interests and concerns. In this way, Huntington posited that the American military would be up to the task of defeating the very real threats it would face, especially during the Cold War context in which Huntington wrote.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics. It would be too burdensome to unpack all of those here, but for clarity, several quick counterarguments to Huntington are worth highlighting. Eliot Cohen, in his work, _Supreme Command,_ argues that the very concept of separate spheres between political and military leaders is flawed. Pointing to leaders such as Lincoln, Clemenceau, and Churchill, and the actions that each took to ensure that the military operated in accordance with each's wishes, Cohen argues that political leaders are ultimately responsible for the outcomes of war, and therefore, are not obligated to grant the military any sort of distinct operating space [@cohen_supreme_2003]. Another scholar, Peter Feaver, argues in his work, _Armed Servants: Agency, Oversight, and Civil-Military Relations_ that a principal-agent approach most accurately conveys how civil-military relations work [@feaver_armed_2005]. American elected civilian leaders, as agents, so argues Huntington, have the responsibility to select military leaders as agents, and to oversee ensuing problems of agency that arise, such as delegating, monitoring, and ensuring the compliance of the military as agents [@feaver_armed_2005]. The point worth highlighting here is that Cohen and Feaver argue that civilian elected leaders have the right to encroach onto what Huntington would call for, a separate military sphere. Note, however, that none of these scholars argue for the opposite - that military leaders should in any way encroach onto or towards the political space.

It is not only in the academic literature that the tradition and worth of an apolitical military have been cultivated. Indeed, the operational military - those who prepare for fighting and dying on the battlefield - have also been central in upholding the military's apolitical traditions. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific who is also well known for the controversy surrounding his conduct in the Korean War and his eventual dismissal by President Truman, told the cadets at West Point in May of 1962.

@@

Current military doctrine and practices likewise reinforce the notion of the military remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General (Retired) Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. General Martin Dempsey, in an opinion piece published in the Washington Post on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine General John Allen for participating in the Republican and Democratic conventions, respectively, stating that "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

It is important to keep in mind, however, that although the notion of an apolitical military is the theoretical goal of the institution, the practical record remains far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory from time to time. What we have established, though, is clear: an apolitical military that performs its military mission well is what enables the American public to put trust and confidence in the military institution. Having examined the nature of the norm of the military as an apolitical institution, let us now briefly examine the topics of the media, public opinion, and polarization. Each of these topics intersects with the norm of the military as an apolitical institution in a unique way.

@@

To test whether there has been variation in opinion-editorial usage by retried military officers in the United States, I will examine opinion-editorial pieces published in the four leading newspapers of the United States since January 1, 2000: the _Wall Street Journal_ (WSJ), the _New York Times_ (NYT), the _Washington Post_ (WaPo), and the _USA Today_. This date range is important for several reasons.

First, let us consider some of the significant ways in which the military as an institution has changed in the past twenty years. Most notably, the post-9/11 US military is a battle-hardened military. The wars in Afghanistan and Iraq continue to the present day, and today's recently retired military generals and admirals are, in many cases, veterans of these conflicts several times over. Moreover, both of these conflicts have thus far failed to reach conclusive ends. Heidi Urben found in a 2009 study that the proportion of military officers who felt that criticizing civilian administrations constituted proper conduct had increased since 1998, reasoning that the likely reason for the change was the frustration encountered while serving in Iraq and Afghanistan [@urben_party_2013].

Moreover, the military since 2000 has undergone significant social changes. The repeal of the "Don't Ask, Don't Tell" Policy, which was instituted under President Clinton in 1992 and prevented homosexual members of the Armed Forces from revealing their sexual preference, was repealed in 2010. Then, in 2015, under the Obama Administration, the Department of Defense opened up all military occupational specialties to women, many of which, particularly those in the combat arms and special forces, had previously been closed off to  women. These changes in the military are just a subset of the profound change that America as a nation has experienced in the past 20 years.

Likewise, the media has also undergone significant changes since 2000. Since the mid-2000s, social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, even as many of the leading newspapers have taken up digital offerings, but overall circulation is still down compared to previous era even after factoring in the digital options available. Estimated total daily newspaper circulation in the US in 2000 was 56 million and 59 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million. Despite these trends, however, the four newspapers that I examine in this project are the largest of all print newspapers in the United States.

Moreover, while there a multitude of ways for military elites to express their opinions and to influence public opinion, such as appearing participating in an interview on television or on the radio, published opinion pieces, especially those that run in a major newspaper, possess certain qualities. First, such types of opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. Second, publishing opinion pieces, especially for major newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by heads of state, members of congress, governmental figures, and heads of industry and academia. So, even though overall newspaper consumption may in fact be down over the past twenty years, there is no evidence that opinion pieces have dropped in popularity.

Attempting to reconcile the facts that the military prides itself on being a non-partisan and apolitical institution, that military officers with experience in Iraq and Afghanistan are more willing to criticize civilian leaders over the conduct of war, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 20 years, and that the US political environment is increasingly polarized along partisan lines, I adopt the following hypotheses:

_$H_1$: The majority of op-eds published by retired military officers will be apolitical in tone._

_$H_2$: There will be an increase in the number of op-eds that criticize civilian administration handling regarding the conduct of war in Iraq and Afghanistan._

@@

I first assemble the data by creating a data set that consists of opinion-editorial pieces published by retired military officers in the four major newspapers previously identified since January 1, 2000. I employ a search using the Factiva Database and condition my search on several filters:

  * sources (WSJ, NYT, WaPo, and USA Today)
  * date range (January 1, 2000 - April 16, 2020)
  * article type (I select editorial, opinion, and commentary)
  * perhaps the most difficult filter item, free text that identifies retired military officers as the   authors. I include terms such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "general," "admiral," and so forth. This is done in order to capture the biographic description that usually accompanies an editorial or opinion piece, such as "General David Petraeus is a retired Army general and former commander of all forces in Afghanistan."

Factiva allows for the option of selecting print only or an "all-sources" option in the source filter. I select the "all-sources" option because I want to capture any online editorials or opinion pieces, in addition to those that are ran in print editions only. Though digital offerings have certainly expanded in the past twenty years, including digital offerings will enable a better comparison of the data over the past twenty years given that overall newspaper circulation is down over the same time period. Factiva also checks for duplicate articles that are published in different versions of the same newspaper, such as the New York Times and the International version of the New York Times. When the same articles appeared in two different versions of the same newspaper, I counted the article only one time.

@@

```{r summary, include = FALSE}
# Read in the Data
dataset <- read_csv(here("data", "MilEditorials.csv"))
dataset %>% print()
# Replace NA with 0 in the data for variables PolicySecurity thru PolAdminName
adjdataset <- dataset %>%
  mutate_at(
    .vars = vars(PolicySecurity:PolAdminName),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

@@

# count the number of observations by type of binary variable and store as new dataframe
countbytype  <- adjdataset %>% count(PolicySecurity, PersTroopSpt, GenAdvice, PolicySocial, Party, PolAdminName)

@@

Table 1 reveals that an overwhelming majority (186 of 216, or 86%) of the total number of opinion pieces written since 2000 in the four major newspapers examined can be classified according to one of the following three categories: **PolicySecurity**, **PersTroopSupport**, and **GenAdvice**. The remaining 30 publications (14% of the total) fall into one of the other three categories: **PolicySocial**, **Party**, and **PolAdminName**.

I consider these 186 observations as being in alignment with traditional civil-military norms on the basis that they address topics which American citizens can reasonably expect retired military officers to address. In other words, I consider retired military officers writing on topics related to security policy, support for members of the Armed Forces, and general advice of a strategic nature not to be problematic. In line with Zaller's framework of elite discourse helping to frame public opinion, these 186 observations demonstrate military elites writing on areas of policy in which they reasonably possess considerable expertise. The remaining 30 observations - those that have been classified as **PolicySocial**, **Party**, or **PolAdminName**, require further examination. There may be nothing inappropriate from the perspective of traditional civil-military norms about these remaining observations, but they have been flagged on the basis of the topics that they address.

>>>>

---
title: "When the Pen Becomes a Sword: Retired Military Officers and Opinion-Editorial Usage in the US"
author: "anonymous"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper seeks to explain how retired military officers have attempted to shape US public opinion through the publication of op-ed pieces in major newspapers. I construct and analysze a data set that captures more than 300 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1975, and ultimately find two trends that stand out. The first is that retired military officers are increasingly writing about topics that are outside of traditional security related topics. The second trend is that retired military officers are increasingly criticizing Presidential administrations, especially that of President Donald Trump, for reasons other than disputes regarding international uses of force. Given the polarizing political climate in the United States and that a Presidential election will occur this fall, these two trends are concerning from the perspective of civil-military relations and suggest that the norm of a professional, apolitical military establishment is under significant strain._

@@

On June 1 2020, the President of the United States, Donald Trump, after giving remarks at the Rose Garden of the White House in Washington D.C., walked to St. John's Church after security forces visibly displaced throngs of protestors in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and after several minutes, returned to the White House.

In the aftermath of what the American mainstream media alleged had amounted to nothing more than a publicity stunt, several members of the President's cabinet members and advisors sought to distance themselves from the President's actions, and more broadly, the President's threat of invoking the Insurrection Act of 1807 to use active-duty forces to quell domestic rioting. The Secretary of Defense, Dr. Mark Esper, stated that he "did not support" use of the 1807 Insurrection Act to deploy active duty military to establish order, and General Mark Milley, the Chairman of the Joint Chiefs of Staff, later told a graduating class from the National Defense University that he "should not have been [joining the President on the walk to St. John's Church]," and that his presence "created a perception of the military involved in domestic politics" [@esper_secretary_2020; @milley_official_2020].

Multiple retired American generals also took to the press to denounce the Trump administration's responce to ongoing protests and riots occurring the aftermath of George Floyd's death in Minneapolis. But for those paying close attention to these retired military elites, it is clear that not all who spoke out did so uniformly. Across a spectrum, there were some retired officers who acknowledged that what had occurred was, at least in a general sense, negative in terms of its impact on civil military relations, while there were also several other officers who went farther in their criticism, directly blasting the President or members of his cabinet for actions taken.

This paper examines the way in which a narrow subset of elites - retired military officers - has attempted to influence public opinion in contemporary America through the use of opinion-editorial pieces. It is an interesting question worthy of examination for several reasons and from a variety of perspectives. First, even after leaving active duty service, military officers remain part of an institution - the military - that maintains considerable trust among the American public. A 2018 Pew Research Center poll determined that 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. Therefore, when leaders of the most trusted institution in the country speak, the public seems to trust them to a high degree. Second, it is an interesting question because as an institution, the military prides itself on being apolitical. In an era of increasing political polarization [@layman_party_2006; @hare_polarization_2014], examining the extent to which opinion-editorial usage by retired military officers has changed may reveal one way in which the military institution has altered its behavior in a polarized political environment, and especially, in the months leading up to the Presidential Election in November of this year.

This paper proceeds in three parts. First, I briefly review the basis upon which the norm of apolitical behavior of the military as an institution is based, as well as describe how a polarized environment may impact these norms. Here I review some of the academic literature on civil-military relations and public opinion, as well as what current military doctrine states on the topic. Second, I describe the analytical approach and the data used in this project. Assembling an original data set that includes more than 300 opinion pieces published by retired military officers in major American newspapers since 1975, I conduct a mixed-methods approach which ultimately reveals two important findings: the first is that although the vast majority of opinion pieces published in major newspapers seem to reflect the military as a partisan institution, retired military officers are increasingly writing about topics that are not immediately related to national security, and second, retired military officers are criticizing if not altogether attacking political administrations more often than they did in previous eras. Third, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a highly polarized political climate.

# The Military as an Apolitical Institution

As a norm, the military acting as an objective, apolitical institution is deep and storied in the United States. In a thriving democracy such as the US, an apolitical military can best serve the Republic by defending it regardless of which political party of group holds power. One of the strongest origins of the norm of the military as an apolitical institution stems from the work of the political scientist, Samuel Huntington. Written in 1957 at the height of the Cold War, his book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced [@huntington_soldier_1957]. The second force was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society, and that these forces were at odds with each other. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution ensured that the military remained "professional," and thus, untainted by political interests and concerns. In this way, Huntington posited that the American military would be up to the task of defeating the very real threats it would face while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant today, came from S.E. Finer, whose 1962 book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962, 22]. Finer's main concern was that of military professionals becoming so professional that they develop the idea that they somehow know better than their elected civilian leadership, particularly on thorny and controversial concerns. Indeed, Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. More important than viewing themselves as professionals, Finer argued that what the military needed in order to stay out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the concept are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situation should be rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military have been cultivated. The operational military - those who prepare for fighting and dying on the battlefield - have also been central in upholding the military's non-partisan traditions. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific who is also well known for the controversy surrounding his conduct in the Korean War and his eventual dismissal by President Truman, told the cadets at West Point in May of 1962.

@@

Current military doctrine and practices likewise reinforce the notion of the military remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General (Retired) Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. General Martin Dempsey, in an opinion piece published in the Washington Post on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although the notion of an apolitical military is the theoretical goal of the institution, the practical record remains far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory from time to time. What we have established, though, is clear: a non-partisan military that performs its military mission well has long been the goal of both civil-military relations theorists and what enables the American public to put trust and confidence in the military as an institution.

@@

To test whether there has been variation in opinion-editorial usage by retired military officers in the United States, I will examine opinion-editorial pieces published in the five leading newspapers of the United States since January 1, 1979: _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. This date range is important, interesting, and rich, as it captures a multitude of changes and debates that have involved the military in this time period, all of which are captured in the dataset constructed for this project.

Consider first that in terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, and a host of smaller but important military efforts in places such as Somalia, Haiti, and Grenada. Each of these conflicts and wars have themselves varied significantly, including in intensity and and their conclusivity.

Second, the military has undergone significant social and institutional changes in this time period. In the late 1970s, the military had ended the draft only somewhat recently (1973). The data includes discussion of institutional culture that rocked the Navy following the Tailhook Scandalin 1991, as well as debate over first instituting and ultimately repealing the "Don't Ask, Don't Tell" Policy. The data also includes discussions of changes occurring in the Obama Administration, such as the opening of military occupational specialties to women, and frustrations expressed by officers at much of President Trump's foreign policy.

Third, the media itself has undergone significant changes since the late 1970s. Since the mid-2000s, social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, but overall circulation is still down today compared to earlier years, even after factoring in expanse of digital options available today. As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million and 54 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million.

Given these significant changes involving the American military over the past forty years, and the substantial changes that have impacted the media and the manner in which Americans consume information, several specific observations are worth noting about opinion-editorial pieces. The first is that opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. They are different from interviews or commentary that may be asked because the author must take time to put forth his or her position in writing. Second, publishing opinion pieces, especially for major newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by figures that include heads of state, members of congress, governmental figures, and leaders of industry and academia.

Attempting to reconcile the facts that the military prides itself on being a non-partisan and apolitical institution, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 40 years, and that the US political environment is increasingly polarized along partisan lines, I adopt the following hypotheses:

_$H_1$: The majority of op-eds published by retired military officers will be apolitical and non-partisan in tone, but there will be an increase in the number of op-eds criticizing civilian administrations in the past several years, as polarization has dramatically worsened._

_$H_2$: Because overall newspaper consumption has dropped in the past 40 years, there will be fewer opinion pieces over time authored by military officers._

@@

To test my hyptheses, I first create a data set that includes all identifiable opinion-editorial pieces authored by retired military officers from January 1, 1979 through June 30, 2020. Using the Factiva and Nexus-Uni databases, I filter a search that utilizes appropriate parameters and uses specific keywords in the byline (author) field such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "general," "admiral," and so forth. I also include in my search both print and online offerings to ensure that I am capturing any opinion pieces published digitally.

@@

```{r summary, include = FALSE}
# Read in the Data
dataset <- read_csv(here("data", "MilEditorials24June.csv"))
dataset %>% print()
# Replace NA with 0 in the data for variables PolicySecurity thru PolAdminName
adjdataset <- dataset %>%
  mutate_at(
    .vars = vars(Warfighting_Ops:Admin_Fig_Name),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

@@

# count the number of observations by type of binary variable and store as new dataframe
countbytype  <- adjdataset %>% count(Warfighting_Ops, TroopSpt_Culture, Strat_Advice, MilPolicy_Social, Domestic_Policy, Admin_Fig_Name) 
countbytype

@@

Table 1 reveals that an overwhelming majority (289 of 334, or 86.5%) of the total number of opinion pieces written since 1979 in the major newspapers examined can be classified according to one of the following three categories: **Warfighting_Ops**, **TroopSpt_Culture**, and **Strat_Advice**. It is also entirely reasonable to conclude that these op-eds written are appropriate from the perspective of civil-military relations in that they address topics which American citizens can reasonably expect retired military officers to address and to bring their expertise to bear upon. In line with Zaller's framework of elite discourse helping to frame public opinion of substantive issue areas, the fact is that the overwhelming majority of opinion articles penned by retired military officers are offered on topics that military expertise can reasonably address based on the expertise they have acquired throughout their careers.

This is not to say that these pieces are entirely without controversy, or to suggest that opinion pieces authored by retired military ought to be free. A review of the data clearly shows, on multiple occasions, retired military officers passionately arguing for the adoption of certain courses of action through the American press. For example,....

The remaining 45 opinion pieces (13.5% of the total) fall into one of the other three categories: **MilPolicy_Social**, **Domestic_Policy**, and **Admin_Fig_Name**. 
The remaining 30 observations - those that have been classified as **PolicySocial**, **Party**, or **PolAdminName**, require further examination. There may be nothing inappropriate from the perspective of traditional civil-military norms about these remaining observations, but they have been flagged on the basis of the topics that they address.
